---
output: html_document
---
#Peruvian Tern Archival Light-level geolocator Analysis
*Analysis conducted by M.T.Hallworth*

Archival light-level geolocators (hereafter geolocators)have been used to track individuals since the early 1990s but were restricted to large organisms because of their large size. Recently, with the miniturization of geolocators, researchers are now able to deploy geolocators on smaller and smaller species. Geolocators are devices that record ambient light levels every 2, 5, or 10 min depending on the model. Geolocators are attached to individuals which then migrate with the device while it records ambient light-levels throughout the year. Once recovered, the data are downloaded and analyzed to determine the time of sunrise and sunset. From the sunrise/sunset data the time of relative noon and midnight are determined. Geographical cooridnates are then derived from the relative ‘noon’ and ‘midnight’ times to give an approximate location of where the individual was throughout the year. The following document outlines the steps for analyzing data from geolocators deployed on Peruvian Terns in northern Chile. 

### Install required packages 

```{r eval = FALSE}
# Check to make sure the required packages are installed on your machine
# If not, they will be installed

reqPackages <- c("devtools","raster","sp","maptools","geoshpere","rgeos","MASS")
get.packages <- reqPackages[!(reqPackages %in% installed.packages()[,"Package"])]
if(length(get.packages)>0) install.packages(get.packages, dependencies=TRUE)

# Install necessary packages from Github using the devtools library #
library(devtools)
install_github("SWotherspoon/SGAT")
install_github("SLisovski/TwGeos")
```

```{r echo = FALSE}
.libPaths("Packages")
```

### Load the required packages
```{r, warning = FALSE, message = FALSE}
library(raster)
library(sp)
library(rgeos)
library(geosphere)
library(SGAT)
library(TwGeos)
library(MASS)
library(maptools)
```

### Read in spatial data 

The following code reads in spatial data for the Americas and the Birdlife International's shapefile for the Peruvivan Terns distribution. For the analysis I added a buffer of 1500km to the current distribution just in case some of the birds use areas outside of the current 'known' distribution. The buffered distribution is used a prior in the light-level analysis. The areas within the dotted lines (see below) have a higher probability than those outside the dotted line. However, locations outside the dotted line are not impossible. 

```{r spatial data, message = FALSE}
# read in a simple world map from the maptools package #
Americas<-raster::shapefile("Spatial_Layers/Americas.shp")
PETEdist<-raster::shapefile("Spatial_Layers/Sternula_lorata.shp")

WGS84<-"+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

laea <- "+proj=laea +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs"

PETEdistB <- rgeos::gBuffer(spTransform(PETEdist,CRS(laea)),byid = TRUE, width = 1500000)
PETEdistBuffer <- spTransform(PETEdistB, CRS(WGS84))
```

```{r echo = FALSE}
plot(gUnaryUnion(PETEdistBuffer), lty = 2)
plot(Americas, add = TRUE, col = "gray88")
plot(PETEdist, add = TRUE, col = "red", border = "red")
plot(gUnaryUnion(PETEdistBuffer), add = TRUE,lty = 2)
```    

    
### Reading in the light data 
```{r}
## ------------------------------------------------------------------------
TernFiles <- list.files("Data",pattern = ".lig",full.names=TRUE)

# Read just the file names for Bird ID
BirdId <- list.files("Data", pattern=".lig")

# Determine the number of birds
nBirds<-length(BirdId)


# Read all files in as Lig files #

Terndata <- lapply(TernFiles,readLig,skip = 1) 

# Keep only the relevant data 

for(i in 1:nBirds){
Terndata[[i]] <- Terndata[[i]][,c(2,4)]
}
```

Here I subset the data to include only the dates when the tag was deployed on the birds
```{r}
## Subset the data to only include when on the birds ##
tern008 <- as.POSIXct(c("2015-09-11","2016-09-02"),format = "%Y-%m-%d",tz = "GMT")
tern009 <- as.POSIXct(c("2015-09-11","2016-09-03"),format = "%Y-%m-%d",tz = "GMT")
tern010 <- as.POSIXct(c("2015-09-12","2016-09-06"),format = "%Y-%m-%d",tz = "GMT")

Terndata[[1]] <- subset(Terndata[[1]],Date > tern008[1] & Date < tern008[2])
Terndata[[2]] <- subset(Terndata[[2]],Date > tern009[1] & Date < tern009[2])
Terndata[[3]] <- subset(Terndata[[3]],Date > tern010[1] & Date < tern010[2])
```

Set the capture location for each individual - c(Long,Lat)
```{r}
# Set the capture coordinates for each bird #

CapLocs<-array(NA,c(nBirds,2))

CapLocs[1,] <- c(-70.357,-23.0544)    # 008
CapLocs[2,] <- c(-70.357,-23.0543)    # 009
CapLocs[3,] <- c(-70.359,-23.0474)    # 010
```

Here are what the data look like - 

```{r echo = FALSE, dpi = 600, fig.width = 8, fig.height = 6, fig.cap = "Figure 1. Raw Light-level data shown throughout the year. Light areas in the figure indicate times when the geolocator recorded light values. Dark areas are times when there was no light or the sensor was covered."}
par(mfrow = c(3,1),mar = c(4,2,0,0))
for(i in 1:nBirds){
lightImage(Terndata[[i]],
           offset = 19,
           zlim = c(0,5))
}
```

### Assigning Twilight events 

```{r}
# Make empty list to store twilight events
twl <- vector('list',nBirds)

seed <- as.POSIXct(rep("2016-01-01 04:00:00",3),format = "%Y-%m-%d %H:%M:%S", tz = "GMT")

for(i in 1:nBirds){
twl[[i]] <- findTwilights(tagdata = Terndata[[i]],
                        threshold = 1,
                        include = seed[i],
                        dark.min = 240) # 4 hours minimum dark period
}
```

Below is the code indicating how I objectively edited the twilight events
```{r}
### ---- Edit twl ---- ###
twlEdit <- vector('list',nBirds)

for(i in 1:nBirds){
twlEdit[[i]] <- twilightEdit(twilights = twl[[i]], 
                    window = 4,           # two days before and two days after
                    outlier.mins = 35,    # difference in mins
                    stationary.mins = 25, # are the other surrounding twilights within 25 mins of one another
                    plot = TRUE)
}
```

Adjust for the recording interval and remove the deleted transitions.

```{r}

 for(i in 1:nBirds){
   twlEdit[[i]]<-twilightAdjust(twilights=twlEdit[[i]], interval=120)
 }

head(twlEdit[[1]])

twlEdit <- lapply(twlEdit,subset,Deleted == FALSE)
```

Save the twilight files for future reference. 
```{r eval = FALSE}

for(i in 1:nBirds){
saveRDS(twlEdit[[i]],paste0("Twilights/",BirdId[[i]],".rds")) # Saves rds file with the BirdId.rds as the name
write.csv(twlEdit[[i]],paste0("Twilights/",BirdId[[i]],".csv"),row.names = FALSE)
}
```

### Calibration
```{r}
# Create a vector with the dates known to be at deployment #
calibration.dates <- vector('list',nBirds)

for(i in 1:nBirds){
calibration.dates[[i]] <- c(Terndata[[i]][1,1],as.POSIXct("2015-09-18",tz="GMT"))
}
```

Extract twilight information during the calibration period
```{r}
# Extract twilight data during calibration period
calibration.data<-vector('list',nBirds)

for(i in 1:nBirds){
  calibration.data[[i]]<-subset(twlEdit[[i]],twlEdit[[i]]$Twilight>=calibration.dates[[i]][1] & 
                                             twlEdit[[i]]$Twilight<=calibration.dates[[i]][2])
}
```

### Determine the zenith angle
```{r}
# create empty vectors to store data #
sun<-z<-zenith0<-zenith1<-twl_t<-twl_deviation<-alpha<-fitml<-vector("list",nBirds)

# loop through each of the two individuals #
for(i in 1:nBirds){
  
  # Calculate solar time from calibration data 
  sun[[i]]  <- solar(calibration.data[[i]][,1])
  
  # Adjust the solar zenith angle for atmospheric refraction
  z[[i]]<- refracted(zenith(sun = sun[[i]],
                            lon = CapLocs[i,1],
                            lat = CapLocs[i,2]))
  
  twl_t[[i]]   <- twilight(tm = calibration.data[[i]][,1], 
                           lon = CapLocs[i,1],
                           lat = CapLocs[i,2],
                           rise = calibration.data[[i]][,2],
                           zenith = quantile(z[[i]],probs=0.5))
  
  # Determine the difference in minutes from when the sun rose and the geolocator said it rose 
  twl_deviation[[i]] <- ifelse(calibration.data[[i]]$Rise, as.numeric(difftime(calibration.data[[i]][,1], twl_t[[i]], units = "mins")),
                         as.numeric(difftime(twl_t[[i]], calibration.data[[i]][,1], units = "mins")))
  
  twl_deviation[[i]]<-subset(twl_deviation[[i]],twl_deviation[[i]]>=0)
  
  # Describe the distribution of the error 
  fitml[[i]] <- fitdistr(twl_deviation[[i]], "log-Normal")
  # save the Twilight model parameters
  alpha[[i]] <- c(fitml[[i]]$estimate[1], fitml[[i]]$estimate[2]) 
}
```

```{r echo = FALSE}
meanAlpha1<-mean(c(alpha[[1]][1],alpha[[2]][1],mean(alpha[[3]][1])))
meanAlpha2<-mean(c(alpha[[1]][2],alpha[[2]][2],mean(alpha[[3]][2])))

ALPHA<-alpha[[1]]
ALPHA[1]<-meanAlpha1
ALPHA[2]<-meanAlpha2
```

```{r}
zenith0 <- zenith1 <- rep(NA,nBirds)

# loop through the birds #
for(i in 1:nBirds){
  # Store the zenith (sun-elevation angle)
  zenith0[i] <-quantile(z[[i]],prob=0.5)
  zenith1[i]<-quantile(z[[i]],prob=0.95)
}
```

```{r, echo = FALSE, fig.cap = ("**Figure 2** *Left* The deviation in twilights from the true twilight at the capture site. *Right* The mean (point estimate) and 95% CI for the zenith angle at the capture location.")}

b<-unlist(twl_deviation)
cols<-c("red","blue","green","yellow","orange","purple","brown","gray","black","pink")
seq <- seq(0,60, length = 100)
par(mfrow=c(1,2),mar=c(4,4,0,0))
hist(b, freq = F,
     yaxt="n",
     ylim = c(0, 0.15),
     xlim = c(0, 60),
     breaks=15,
     col="gray",
     main = "",
     xlab = "Twilight error (mins)")
axis(2,las=2)
for(i in 1:nBirds){
lines(seq, dlnorm(seq, alpha[[i]][1], alpha[[i]][2]), col = cols[i], lwd = 3, lty = 2)
}

#Zenith angle plot
par(bty="l")
plot(median(z[[1]],na.rm=TRUE),xlim=c(1,3.5),ylim=c(85,100),pch=19,ylab="Zenith Angle",xlab="PETU",col=cols[1])
segments(1,quantile(z[[1]],probs=0.025),1,quantile(z[[1]],probs=0.975),col=cols[1])
for(i in 2:nBirds){
  par(new = TRUE)
  plot(median(z[[i]],na.rm=TRUE)~i,xlim=c(1,3.5),ylim=c(85,100),pch=19,yaxt="n",xaxt="n",ylab="",xlab="",col=cols[i])
  segments(i,quantile(z[[i]],probs=0.025),i,quantile(z[[i]],probs=0.975),col=cols[i])
}
``` 

### Determine the initial path 

This step uses a simple threshold method to determine the initial path to parameterize the MCMC model. See below. 
```{r warning = FALSE}
# Create empty vectors to store objects #
path<-vector('list',nBirds)

tolvalues <- array(NA,c(nBirds,3))
tolvalues[,1]<-0.1
tolvalues[,2]<-0
tolvalues[,3]<-0.1

for(i in 1:nBirds){  
   path[[i]] <- thresholdPath(twilight = twlEdit[[i]]$Twilight,
                             rise = twlEdit[[i]]$Rise,
                             zenith = zenith1[i],
                             tol = tolvalues[i,])
}
```

Extract the initial locations
```{r}
x0 <- z0 <- vector('list',nBirds)

for(i in 1:nBirds){
  # Take the location estimates created above
x0[[i]]<- path[[i]]$x

  # the model also needs the mid-points - generate those here
z0[[i]]<- trackMidpts(x0[[i]])
}
```

Here we set a realistic flight parameter.
```{r}
beta <- c(0.7, 0.08)
```

Here we set the boundaries of the possible locations where the Peruvian Tern may occur over the time that the tag was deployed. 

```{r}
# set xlim and ylim values need to span the range of your dataset
xlim <- c(-100, -60)
ylim <- c(-90, 90)
```

Here we define the function to create the location prior (ie, distribution buffer)
```{r}
## Function to construct a land/sea mask
distribution.mask <- function(xlim, ylim, res = c(0.25,0.25), land = TRUE, shape) {
    r <- raster(res = res,
                xmn = xlim[1], 
                xmx = xlim[2], 
                ymn = ylim[1], 
                ymx = ylim[2], 
                crs = proj4string(shape))

    r <- cover(rasterize(shape, shift = c(-360, 0), r, 1, silent = TRUE), 
        rasterize(shape, r, 1, silent = TRUE), rasterize(elide(shape, 
           shift = c(360, 0)), r, 1, silent = TRUE))
    r <- as.matrix(is.na(r))[nrow(r):1, ]
   if (land) 
        r <- !r
   xbin <- seq(xlim[1], xlim[2], length = ncol(r) + 1)
    ybin <- seq(ylim[1], ylim[2], length = nrow(r) + 1)

    function(p) {
        r[cbind(.bincode(p[, 2], ybin), .bincode(p[, 1], xbin))]
    }
}
```

```{r echo = FALSE}
crs(Americas)<-WGS84
crs(PETEdist)<-WGS84
crs(PETEdistBuffer) <- WGS84
```

Use the function to create the prior distribution surface using the 1500km buffer around the current 'known' distribution. 
```{r}
 is.dist <- distribution.mask(shape=PETEdistBuffer,
                             xlim = xlim,
                             ylim = ylim,
                             res = c(0.25,0.25), # 0.25 x 0.25 degree resolution
                             land = TRUE)

# Define the log prior for x and z
log.prior <- function(p) {
    f <- is.dist(p)
    ifelse(f | is.na(f), 0, -10)
}
```

Here we set the time when birds are stationary. These are individual twilight events. Generally there are two twilights per day. Therefore, 60 twilights represents approximately a 30 day window. 
```{r}
# set time when birds are stationary 
fixedx <- vector('list',nBirds)
for(i in 1:nBirds){
fixedx[[i]] <- rep(FALSE, length(twlEdit[[i]]$Twilight))
fixedx[[i]][1:60]<-TRUE
fixedx[[i]][(length(fixedx[[i]])-60):length(fixedx[[i]])]<-TRUE
x0[[i]][fixedx[[i]],1] <- CapLocs[i,1]
x0[[i]][fixedx[[i]],2] <- CapLocs[i,2]
}
```

### The MCMC model
```{r}
model <-  vector('list', nBirds)

for(i in 1:nBirds){
model[[i]]<- thresholdModel(twilight = twlEdit[[i]]$Twilight,
                            rise = twlEdit[[i]]$Rise,
                            twilight.model = "ModifiedLogNormal",
                            alpha = alpha[[i]],
                            beta = beta,
                            # Here is where we set the constraints for land
                            logp.x = log.prior, 
                            logp.z = log.prior, 
                            x0 = x0[[i]],
                            z0 = z0[[i]],
                            fixedx = fixedx[[i]],
                            zenith = zenith1[i])
}
```


This defines the error distribution around each location
```{r}
proposal.x <- proposal.z <- vector('list',nBirds)

for(i in 1:nBirds){
proposal.x[[i]] <- mvnorm(S=diag(c(0.0025,0.0025)),n=nlocation(x0[[i]]))
proposal.z[[i]] <- mvnorm(S=diag(c(0.0025,0.0025)),n=nlocation(z0[[i]]))
}
```

Here we run the model
```{r eval = FALSE}
a <- Sys.time()
 fit <- xsum <- zsum <- vector('list', nBirds)
 
 for(i in 1:nBirds){
 fit[[i]] <- estelleMetropolis(model = model[[i]],
                               proposal.x = proposal.x[[i]],
                               proposal.z = proposal.z[[i]],
                               iters = 2000, # This value sets the number of iterations to run
                               thin = 10,
                               chains = 3)

 xsum[[i]] <- locationSummary(fit[[i]]$x,collapse = TRUE)
 zsum[[i]] <- locationSummary(fit[[i]]$z,collapse = TRUE)

proposal.x[[i]] <- mvnorm(S=diag(c(0.0025,0.0025)),n=nlocation(cbind(xsum[[i]]$'Lon.50%',xsum[[i]]$'Lat.50%')))
proposal.z[[i]] <- mvnorm(S=diag(c(0.0025,0.0025)),n=nlocation(cbind(zsum[[i]]$'Lon.50%',zsum[[i]]$'Lat.50%')))

 fit[[i]] <- estelleMetropolis(model = model[[i]],
                              proposal.x = proposal.x[[i]],
                               proposal.z = proposal.z[[i]],
                               x0 = cbind(xsum[[i]]$'Lon.50%',xsum[[i]]$'Lat.50%'),
                               z0 = cbind(zsum[[i]]$'Lon.50%',zsum[[i]]$'Lat.50%'),
                               iters=2000, # This value sets the number of iterations to run
                               thin=10,
                               chains=3)
 
# Final Run
  xsum[[i]] <- locationSummary(fit[[i]]$x,collapse = TRUE)
  zsum[[i]] <- locationSummary(fit[[i]]$z,collapse = TRUE)

 proposal.x[[i]] <- mvnorm(chainCov(fit[[i]]$x),s=0.1)
 proposal.z[[i]] <- mvnorm(chainCov(fit[[i]]$z),s=0.1)
 
 # Note the increase in number of interations - this takes a bit longer to run
fit[[i]] <- estelleMetropolis(model = model[[i]],
                               proposal.x = proposal.x[[i]],
                               proposal.z = proposal.z[[i]],
                               x0=cbind(xsum[[i]]$'Lon.50%',xsum[[i]]$'Lat.50%'),
                               z0=cbind(zsum[[i]]$'Lon.50%',zsum[[i]]$'Lat.50%'),
                               iters=5000,  # This value sets the number of iterations to run
                               thin=10,
                               chains=3)
 }
Sys.time()- a
```

```{r echo = FALSE, eval = FALSE}
saveRDS(fit, paste0("Data/MCMCfit/fit_",format(Sys.Date(),"%b_%d_%Y"),".rds"))
```

```{r echo = FALSE}
fit <- readRDS("Data/MCMCfit/fit_Feb_13_2017.rds")
```

### Summarizing the results 
```{r}
# This step makes an empty raster #
r <- raster(res = c(0.25,0.25),
            xmn=xlim[1],
            xmx=xlim[2],
            ymn=-90,
            ymx=90)
```
```{r}
S <- Sp <- vector('list',nBirds)
for(i in 1:nBirds){
S[[i]] <- slices(type="intermediate",
                 weights = rep(0.5,length(fit[[i]][[1]]$time)),
                 breaks="day",
                 mcmc=fit[[i]],
                 grid=r,
                 include.lowest = FALSE, 
                 right = FALSE)
Sp[[i]] <- slices(type="primary",
                 weights = rep(0.5,length(fit[[i]][[1]]$time)),
                 breaks="day",
                 mcmc=fit[[i]],
                 grid=r,
                 include.lowest = FALSE, 
                 right = FALSE)
}
```

```{r echo = FALSE, message = FALSE}
library(mth, lib.loc = "C:/Users/hallworthm/R_Library")
```

```{r, dpi = 600}
Schedule <- lapply(S, # object
                   mth::MigSchedule, # function
                   prob = 0.95,
                   known.breed = c("2015-09-01","2015-10-01"),
                   known.winter = c("2016-01-01","2016-02-01"),
						       rm.lat.equinox = TRUE,
						       days.omit = 3,
						       progress = FALSE,
						       plot = TRUE,
						       plot.legend = FALSE,
						       latAllow = c(0.5,2))

```

Here we look at the number of shading events within each day. Shading events should go up when they are nesting / incubating. If there two periods when shading events are quite high - it could mean that they are nesting twice. However, there may be many explainations for the shading events (weather, diving, etc.)

```{r, dpi = 600}
shading <- vector('list',nBirds)

for(i in 1:nBirds){
shading[[i]] <- mth::CountShadingEvents(lightdata = Terndata[[i]],
                                        MCMCresult = S[[i]],
                                        threshold = 1,
                                        plot = TRUE)
}
``` 

```{r echo = FALSE, eval = FALSE}
birds <- cols <- vector('list',nBirds)
cols[[1]] <- colorRampPalette(c("darkorchid1","darkorchid4"),alpha = T)
cols[[2]] <- colorRampPalette(c("tomato1","tomato3"),alpha = T)
cols[[3]] <- colorRampPalette(c("springgreen2","springgreen4"),alpha = T)
for(b in 1:nBirds){
KML(Schedule[[b]]$movements, subfolder.name = paste0(substr(BirdId[b],start = 1, stop = 16),".kml"), col = cols[[b]](100),overwrite = TRUE)
}
```
